# Data Import

Let’s face it — getting your data *into* R is half the battle. Once it’s in, the magic begins.

In this section, we’ll walk through how to import real experimental data using the `tidyverse` package. And good news: we’re going to use a dataset that mimics real-life biological experiments specifically designed for this book. No more pretending your experiments look like iris flowers.

---

## Meet Your New Favorite Dataset

We’ll be using a dataset that mimics real-life biological experiments — multiple treatments, time points, and measurements like gene expression, blood pressure, and drug response.

You can download it using this link.  

This dataset originates from a multi-center longitudinal clinical study designed to evaluate the response of 1,200 cancer patients to three experimental drugs: ***Drug A, Drug B, and Drug C*** across five cancer types: lung, breast, colon, prostate, and melanoma. Patients were recruited from 10 different hospitals and followed over six timepoints spanning 12 weeks, with multiple technical replicates per sample.  
The dataset includes patient demographics (age, sex), clinical site information, treatment group, disease type, response status (responder vs. non-responder), longitudinal measurements of two simulated biomarkers, blood pressure, drug dosage, and gene expression levels of two genes.  
The study’s scope is to investigate how these biomarkers and gene expression profiles change over time in relation to drug treatment, dosage, patient characteristics, and clinical response, providing a rich resource for exploring treatment efficacy, molecular correlates of response, and demographic or center-specific effects.

### 🧬 Column Descriptions

| Column           | Description |
|------------------|-------------|
| `PatientID`      | Unique ID for each patient (e.g., "P000123"). Used to track repeated measurements. |
| `Center`         | Clinical site where the patient was enrolled. Used to explore hospital-level variability. |
| `Age`            | Age of the patient in years. Allows age-based subgroup analysis. |
| `Sex`            | Biological sex ("Male" or "Female"). May affect treatment response. |
| `Group`          | Drug group assigned to the patient: "Drug A", "Drug B", or "Drug C". |
| `DiseaseType`    | Type of cancer: "Lung", "Breast", "Colon", "Prostate", or "Melanoma". |
| `ResponseStatus` | Final outcome: "Responder" or "Non-responder" after 12 weeks of treatment. |
| `Timepoint`      | When the sample was collected: "Baseline", "Week1", ..., "Week12". |
| `Biomarker1`     | Simulated concentration of a protein biomarker (e.g., CRP). |
| `Biomarker2`     | Simulated concentration of a cytokine (e.g., IL-6). |
| `Replicate`      | Technical replicate ID (1–5 per patient-timepoint). |
| `BloodPressure`  | Systolic blood pressure in mmHg. Influenced by age and possibly treatment. |
| `Dosage_mg`      | Dose of the drug in milligrams. Varies by group and timepoint. |
| `GeneX_expr`     | Expression of Gene X (log2 scale). Higher in responders. |
| `GeneY_expr`     | Expression of Gene Y (log2 scale). Higher in Lung cancer patients. |

---

We'll use this dataset in most of the examples throughout the book, so feel free to open it in a spreadsheet program to peek at it before we load it.

---

## Load the Tidyverse First

Before we can import the data, we need to load the package that gives us friendly tools to work with data: the `tidyverse`.  

If you don’t have it installed yet, you can do that with:

```{r install_tidyverse, eval=FALSE}
install.packages("tidyverse")
```

After the installation is complete, load the tidyverse package:  

```{r load_tidyverse, eval=FALSE}
library(tidyverse)
```


> 💡 The tidyverse is an umbrella package (i.e a package containing several other packages). It includes readr for reading CSV files, dplyr for data wrangling, and ggplot2 for plotting in addition to a full toolkit for modern R data science.  


## Importing the Dataset


To import the dataset, we’ll use the `read_csv()` function from the readr package (part of the tidyverse). It’s faster and cleaner than the base R `read.csv()` function.


```{r load_dataset, eval=FALSE}
# Load the dataset
bio_data <- read_csv("data/clean_bioexperiment.csv")

# Peek at it
glimpse(bio_data)

```


>Common Gotchas:  
- File not found? Make sure you're working in the right project. Open the .Rproj file before you start coding!  
- Wrong slashes? Always use forward slashes / in file paths — even on Windows.  
- Excel file? We’ll cover .xlsx files later using readxl::read_excel() — for now we’re sticking with .csv for simplicity.  





































































